---
title: "Unlocking single cell spatial omics analyses with scdney"
output:
 rmarkdown::html_document:
    toc: true
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Unlocking single cell spatial omics analyses with scdney}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  #collapse = TRUE,
  #eval = FALSE,
  comment = "#>",
  cache = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{=html}
<style>
.question {
  padding: 1em;
  background: lightcyan;
  color: black;
  border-radius: 10px;
}
</style>
```
**Presenting Authors**

Farhan Ameen$^{1,2,3}$, Alex Qin$^{1,2,3}$, Shreya Rao$^{1,2,3}$, Ellis
Patrick$^{1,2,3}$.

$^1$ Westmead Institute for Medical Research, University of Sydney,
Australia\
$^2$ Sydney Precision Data Science Centre, University of Sydney,
Australia\
$^3$ School of Mathematics and Statistics, University of Sydney,
Australia

<br/> Contact: ellis.patrick\@sydney.edu.au

## Overview

Understanding the interplay between different types of cells and their
immediate environment is critical for understanding the mechanisms of
cells themselves and their function in the context of human diseases.
Recent advances in high dimensional in situ cytometry technologies have
fundamentally revolutionized our ability to observe these complex
cellular relationships providing an unprecedented characterisation of
cellular heterogeneity in a tissue environment.

### Description

In this workshop we will introduce some of the key analytical concepts
needed to analyse data from high dimensional spatial omics technologies
such as, PhenoCycler, IMC, Xenium and MERFISH. We will show how
functionality from our Bioconductor packages simpleSeg, FuseSOM,
scClassify, scHot, spicyR, listClust, statial, scFeatures and ClassifyR
can be used to address various biological hypotheses. By the end of this
workshop attendees will be able to implement and assess some of the key
steps of a spatial analysis pipeline including cell segmentation,
feature normalisation, cell type identification, microenvironment and
cell-state characterisation, spatial hypothesis testing and patient
classification. Understanding these key steps will provide attendees
with the core skills needed to interrogate the comprehensive spatial
information generated by these exciting new technologies.

### Pre-requisites

It is expected that students will have:

-   basic knowledge of R syntax,
-   this workshop will not provide an in-depth description of
    cell-resolution spatial omics technologies.

### *R* / *Bioconductor* packages used

Several single cell R packages will be used from the scdney package, for
more information visit: <https://sydneybiox.github.io/scdney/>.

### Learning objectives

-   Understand and visualise spatial omics datasets.
-   Identify key biological questions that can be addressed with these
    technologies and spatial analysis.
-   Understand the key analytical steps involved in spatial omics
    analysis, and perform these steps using R.
-   Evaluate the performance of data normalisation and cell
    segmentation.
-   Understand and generate individual feature representations from
    spatial omics data.
-   Develop appreciation on how to assess performance of classification
    models.
-   Perform disease outcome prediction using the feature representation
    and robust classification framework.

# Context

In the following we will re-analyse some IMC data [(Ferguson et al, 2022)](https://doi.org/10.1158/1078-0432.CCR-22-1332) profiling the spatial landscape of head and neck cutaneous squamous cell carcinomas (HNcSCC), the second most common type of skin cancer. The majority of HNcSCC can be treated with surgery and good local control, but a subset of large tumours infiltrate subcutaneous tissue and are considered high risk for local recurrence and metastases. The key conclusion of this manuscript (amongst others) is that spatial information about cells and the immune environment can be used to predict primary tumour progression or metastases in patients. We will use our spicy workflow to reach a similar conclusion.

## Workshop

## Setup

### Installation

```{r install, warning=FALSE, message=FALSE, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install(c( "simpleSeg", "cytomapper", "scClassify", "scHOT", "FuseSOM","glmnet", "spicyR", "lisaClust","Statial", "scFeatures", "ClassifyR", "tidyverse", "scater", "SingleCellExperiment", "STexampleData", "SpatialDatasets", "tidySingleCellExperiment", "scuttle", "batchelor"))
```

### Load packages

```{r library, warning=FALSE, message=FALSE}

# packages from scdney
library(SpatialDatasets)
library(simpleSeg)
library(FuseSOM)
library(lisaClust)
library(spicyR)
library(Statial)
library(ClassifyR)

# Other required packages
library(tidySingleCellExperiment)
library(tidyverse)
library(SingleCellExperiment)


theme_set(theme_classic())

nCores <- 8  # Feel free to parallelise things if you have the cores to spare.
BPPARAM <- simpleSeg:::generateBPParam(nCores)
options("restore_SingleCellExperiment_show" = TRUE)

```

### Download images

Feel free to follow along for this section. Much of the segmentation pipeline is computationally demanding, and most likely won't be able to run on the Bioconductor Galaxy servers.

```{r}
# Reads images to cache in a directory specified by the pathToImages variable.
pathToImages <- SpatialDatasets::Ferguson_Images()
tmp <- tempfile() # Creates a temporary directory.
unzip(pathToImages, exdir = tmp) # Unzips files to the temporary directory.

# Store images in a CytoImageList on_disk as h5 files to save memory.
images <- cytomapper::loadImages(
  tmp, # Read files from the temporary directory.
  single_channel = TRUE,
  on_disk = TRUE,
  h5FilesPath = HDF5Array::getHDF5DumpDir(),
  BPPARAM = BPPARAM
)

mcols(images) <- S4Vectors::DataFrame(imageID = names(images))
```

```{r}
# Clean channel names
cn <- channelNames(images) # Read in channel names
head(cn)

cn <- sub(".*_", "", cn) # Remove preceding letters
cn <- sub(".ome", "", cn) # Remove the .ome
head(cn)

channelNames(images) <- cn # Reassign channel names

```

```{r}
# Clean image names
head(names(images))

nam <- sapply(strsplit(names(images), "_"), `[`, 3)
head(nam)

names(images) <- nam # Reassigning image names
mcols(images)[["imageID"]] <- nam # Reassigning image names
```

## SimpleSeg: Segmentation

If your images are stored in a `list` or `CytoImageList` they can be segmented with a simple call to `simpleSeg()`. To summarise, `simpleSeg()` is an R implementation of a simple segmentation technique which traces out the nuclei using a specified channel using `nucleus` then dilates around the traced nuclei by a specified amount using `discSize`. The nucleus can be traced out using either one specified channel, or by using the principal components of all channels most correlated to the specified nuclear channel by setting `pca = TRUE`.

In the particular example below, we have asked `simpleSeg` to do the following: 

By setting `nucleus = c("HH3")`, we've asked simpleSeg to trace out the nuclei signal in the images using the HH3 channel.
By setting `pca = TRUE`, simpleSeg segments out the nuclei mask using a principal component analysis of all channels and using the principal components most aligned with the nuclei channel, in this case, HH3.
By setting `cellBody = "dilate"`, simpleSeg uses a dilation strategy of segmentation, expanding out from the nucleus by a specified `discSize`.
By setting `discSize = 3`, simpleSeg dilates out from the nucleus by 3 pixels.
By setting `sizeSelection = 20`, simpleSeg ensures that only cells with a size greater than 20 pixels will be used.
By setting `transform = "sqrt"`, simpleSeg square root transforms each of the channels prior to segmentation.
By setting `tissue = c("panCK", "CD45", "HH3")`, we specify a tissue mask which simpleSeg uses, filtering out all background noise outside the tissue mask. This is important as these are tumour cores, wand hence circular, so we'd want to ignore background noise which happens outside of the tumour core. 

There are many other parameters that can be specified in simpleSeg (`smooth`, `watershed`, `tolerance`, and `ext`), and we encourage the user to select the best parameters which suit their biological context.

```{r}
masks <- simpleSeg(images,
                   nucleus = c("HH3"),
                   pca = TRUE,
                   cellBody = "dilate",
                   discSize = 3,
                   sizeSelection = 20,
                   transform = "sqrt",
                   tissue = c("panCK", "CD45", "HH3"),
                   cores = nCores
                   )
```

### Visualise outlines

The `plotPixels` function in `cytomapper` makes it easy to overlay the mask on top of the nucleus intensity marker to see how well our segmentation process has performed. Here we can see that the segmentation appears to be performing reasonably.

If you see over or under-segmentation of your images, `discSize` is a key parameter in `simpleSeg()` for optimising the size of the dilation disc after segmenting out the nuclei.

```{r}
plotPixels(image = images["F3"], 
           mask = masks["F3"],
           img_id = "imageID", 
           colour_by = c("HH3"), 
           display = "single",
           colour = list(HH3 = c("black","blue")),
           legend = NULL,
           bcg = list(
             HH3 = c(1, 1, 2)
           ))
```

If you wish to visualise multiple markers instead of just the HH3 marker and see how the segmentation mask performs, this can also be done. Here, we can see that our segmentation mask has done a good job of capturing the CD31 signal, but perhaps not such a good job of capturing the FXIIIA signal, which often lies outside of our dilated nuclear mask. This could suggest that we might need to increase the `discSize` of our dilation.

```{r}
plotPixels(image = images["F3"], 
           mask = masks["F3"],
           img_id = "imageID", 
           colour_by = c("HH3", "CD31", "FX111A"), 
           display = "single",
           colour = list(HH3 = c("black","blue"),
                         CD31 = c("black", "red"),
                         FX111A = c("black", "green") ),
           legend = NULL,
           bcg = list(
             HH3 = c(1, 1, 2),
             CD31 = c(0, 1, 2),
             FX111A = c(0, 1, 1.5)
           ))
```

::: question
**Questions**

1. Is there any information we're not capturing with this segmentation?
2. What parameters might you change to improve the segmentation?
3. What are some intrinsic limitations of the simpleSeg method?
:::

In order to characterise the phenotypes of each of the segmented cells, `measureObjects()` from `cytomapper` will calculate the average intensity of each channel within each cell as well as a few morphological features. By default, the `measureObjects()` function will return a `SingleCellExperiment` object, where the channel intensities are stored in the `counts` assay and the spatial location of each cell is stored in `colData` in the `m.cx` and `m.cy` columns. 

However, you can also specify `measureObjects()` to return a `SpatialExperiment` object by specifying `return_as = "spe"`. As a `SpatialExperiment` object, the spatial location of each cell is stored in the `spatialCoords` slot, as `m.cx` and `m.cy`, which simplifies plotting. In this demonstration, we will return a `SpatialExperiment` object.

```{r}
# Summarise the expression of each marker in each cell
cells <- cytomapper::measureObjects(masks,
                                    images,
                                    img_id = "imageID",
                                    return_as = "spe",
                                    BPPARAM = BPPARAM)

spatialCoordsNames(cells) <- c("x", "y")
```

Next, to associate features in our image with disease progression, it is important to read in information which links image identifiers to their progression status. We will do this here, making sure that our `imageID` column matches.

```{r}
clinical <- read.csv(
  system.file(
    "extdata/clinicalData_TMA1_2021_AF.csv",
    package = "spicyWorkflow"
  )
)

rownames(clinical) <- clinical$imageID
clinical <- clinical[names(images), ]

colData(cells) <- cbind(colData(cells), clinical[cells$imageID, ])
```

## simpleSeg: Normalisation

We've made it easy to follow on from the normalisation section. Our `SpatialDatasets` package has the segmented SCE file stored in the function `spe_Ferguson_2022()`.

```{r data}
#Download data now
spe = SpatialDatasets::spe_Ferguson_2022()

spe$x = spatialCoords(spe)[, "x"]
spe$y = spatialCoords(spe)[, "y"]
```

### Normalise the data

We should check to see if the marker intensities of each cell require some form of transformation or normalisation. The reason we do this is two-fold:  
1) The intensities of images are often highly skewed, preventing any meaningful downstream analysis.  
2) The intensities across different images are often different, meaning that what is considered "positive" can be different across images.

By transforming and normalising the data, we aim to reduce these two effects. Here we extract the intensities from the `counts` assay. Looking at CD3 which should be expressed in the majority of the T cells, the intensities are clearly very skewed, and it is difficult to see what is considered a CD3- cell, and what is a CD3+ cell.

```{r, fig.width=5, fig.height=5}
# Plot densities of CD3 for each image.
cells |> 
  join_features(features = rownames(cells), shape = "wide", assay = "counts") |> 
  ggplot(aes(x = CD3, colour = imageID)) + 
  geom_density() + 
  theme(legend.position = "none")
```

::: question
**Questions**

Can we see what is a CD3+ vs a CD3- cell here?
:::

### Dimension reduction and visualisation

As our data is stored in a `SpatialExperiment` we can also use `scater` to perform and visualise our data in a lower dimension to look for batch effects in our images. We can see that before normalisation, our UMAP shows a clear batch effect between images. 

```{r}
# Usually we specify a subset of the original markers which are informative to separating out distinct cell types for the UMAP and clustering.
ct_markers <- c("podoplanin", "CD13", "CD31",
                "panCK", "CD3", "CD4", "CD8a",
                "CD20", "CD68", "CD16", "CD14", "HLADR", "CD66a")

set.seed(51773)
# Perform dimension reduction using UMAP.
cells <- scater::runUMAP(
  cells,
  subset_row = ct_markers,
  exprs_values = "counts"
)

# Select a subset of images to plot.
someImages <- unique(cells$imageID)[c(1, 5, 10, 20, 30, 40)]

# UMAP by imageID.
scater::plotReducedDim(
  cells[, cells$imageID %in% someImages],
  dimred = "UMAP",
  colour_by = "imageID"
)
```

We can transform and normalise our data using the `normalizeCells` function. In the `normalizeCells()` function, we specify the following parameters.
`transformation` is an optional argument which specifies the function to be applied to the data. We do not apply an arcsinh transformation here, as we already apply a square root transform in the `simpleSeg()` function.
`method = c("trim99", "mean", PC1")` is an optional argument which specifies the normalisation method/s to be performed. Here, we:
1) Trim the 99th percentile
2) Divide by the mean
3) Remove the 1st principal component
`assayIn = "counts"` is a required argument which specifies what the assay you'll be taking the intensity data from is named. In our context, this is called `counts`.

This modified data is then stored in the `norm` assay by default. We can see that this normalised data appears more bimodal, not perfectly, but likely to a sufficient degree for clustering, as we can at least observe a clear CD3+ peak at 1.00, and a CD3- peak at around 0.3.

```{r, fig.width=5, fig.height=5}
# Leave out the nuclei markers from our normalisation process. 
useMarkers <- rownames(cells)[!rownames(cells) %in% c("DNA1", "DNA2", "HH3")]

# Transform and normalise the marker expression of each cell type.
cells <- normalizeCells(cells,
                        markers = useMarkers,
                        transformation = NULL,
                        method = c("trim99", "mean", "PC1"),
                        assayIn = "counts",
                        cores = nCores
)

# Plot densities of CD3 for each image
cells |> 
  join_features(features = rownames(cells), shape = "wide", assay = "norm") |> 
  ggplot(aes(x = CD3, colour = imageID)) + 
  geom_density() + 
  theme(legend.position = "none")
```

::: question
**Questions**

Can we see what is a CD3+ vs a CD3- cell here?
:::

```{r}
set.seed(51773)
# Perform dimension reduction using UMAP.
cells <- scater::runUMAP(
  cells,
  subset_row = ct_markers,
  exprs_values = "norm",
  name = "normUMAP"
)

someImages <- unique(cells$imageID)[c(1, 5, 10, 20, 30, 40)]

# UMAP by imageID.
scater::plotReducedDim(
  cells[, cells$imageID %in% someImages],
  dimred = "normUMAP",
  colour_by = "imageID"
)
```

::: question
**Questions**

1. What are we hoping to achieve with normalisation?
2. What has changed between the pre-normalisation and post-normalisation correction UMAP?
:::

## FuseSOM

To dissect more information about cell interactions with each other, and cell-specific differences between groups, we next perform clustering. The `runFuseSOM()` function from the `FuseSOM` package conveniently runs clustering on any `SingleCellExperiment` object, by specifying the number of clusters under `numClusters`.

```{r FuseSOM}
# Set seed.
set.seed(51773)

# Generate SOM and cluster cells into 10 groups
cells <- runFuseSOM(
  cells,
  markers = ct_markers,
  assay = "norm",
  numClusters = 10
)

```

We can also observe how reasonable our choice of `k = 10` was, using the `estimateNumCluster()` and `optiPlot()` functions. Here we examine the Gap method, but others such as Silhouette and Within Cluster Distance are also available. We can see that there are elbow points in the gap statistic at `k = 7`, `k = 10`, and `k = 11`. We've specified `k = 10`, striking a good balance between the number of clusters and the gap statistic.

```{r}
cells <- estimateNumCluster(cells, kSeq = 2:30)
optiPlot(cells, method = "gap")
```

::: question
**Questions**

1. Was our choice of `k = 10` a good choice?
2. What other `k` might have made sense?
3. From a biological perspective, what would increasing `k` do?
:::

### Attempt to interpret the phenotype of each cluster

We can begin the process of understanding what each of these cell clusters are by using the `plotGroupedHeatmap` function from `scater`. At the least, here we can see we capture all the major immune populations that we expect to see, including the CD4 and CD8 T cells, the CD20+ B cells, the CD68+ myeloid populations, the CD66+ granulocytes, the podoplanin+ epithelial cells, and the panCK+ tumour cells.

```{r}
# Visualise marker expression in each cluster.
scater::plotGroupedHeatmap(
  cells,
  features = ct_markers,
  group = "clusters",
  exprs_values = "norm",
  center = TRUE,
  scale = TRUE,
  zlim = c(-3, 3),
  cluster_rows = FALSE,
  block = "clusters"
)
```

Given domain-specific knowledge of the tumour-immune landscape, we can go ahead and annotate these clusters as cell types given their expression profiles.

```{r}
cells <- cells |>
  mutate(cellType = case_when(
    clusters == "cluster_1" ~ "GC", # Granulocytes
    clusters == "cluster_2" ~ "MC", # Myeloid cells
    clusters == "cluster_3" ~ "SC", # Squamous cells
    clusters == "cluster_4" ~ "EP", # Epithelial cells
    clusters == "cluster_5" ~ "SC", # Squamous cells
    clusters == "cluster_6" ~ "TC_CD4", # CD4 T cells
    clusters == "cluster_7" ~ "BC", # B cells
    clusters == "cluster_8" ~ "EC", # Endothelial cells
    clusters == "cluster_9" ~ "TC_CD8", # CD8 T cells
    clusters == "cluster_10" ~ "DC" # Dendritic cells
  ))
```

We might also be interested in how these cell types are distributed on the images themselves. Here we examine the distribution of clusters on image F3, noting the healthy epithelial and endothelial structures surrounded by tumour cells. 

```{r}
reducedDim(cells, "spatialCoords") <- spatialCoords(cells)

cells |> 
  filter(imageID == "F3") |> 
  plotReducedDim("spatialCoords", colour_by = "cellType")

```

We find it always useful to check the number of cells in each cluster. 

```{r}
# Check cell type frequencies.
cells$cellType |>
  table() |>
  sort()
```

::: question
**Questions**

1. Does our distribution of cell types make sense?
2. What would we normally expect as the most expressed cell type and the least expressed cell type for this dataset?
:::

We can also use the UMAP we computed earlier to visualise our data in a lower dimension and see how well our annotated cell types cluster out.

```{r}
# UMAP by cell type
scater::plotReducedDim(
  cells[, cells$imageID %in% someImages],
  dimred = "normUMAP",
  colour_by = "cellType"
)
```


## spicyR: Test spatial relationships

[spicyR package](https://www.bioconductor.org/packages/release/bioc/html/spicyR.html)

Put simply, spicyR uses the L-function to determine localisation or dispersion between cell types. The L-function is an arbitrary measure of "closeness" between points, with larger values suggesting increased localisation or attraction, and lower values suggesting dispersion or avoidance between cell types.

Here, we quantify spatial relationships using a combination of 10 radii from 10 to 100 by specifying `Rs = 1:10*10` and account for global tissue structure using `sigma = 50`. Further information on how to optimise these parameters can be found in the [vignette](https://bioconductor.org/packages/release/bioc/vignettes/spicyR/inst/doc/spicyR.html) and the spicyR [paper](https://doi.org/10.1093/bioinformatics/btac268).


```{r}
spicyTest <- spicy(cells,
                  condition = "group",
                  cellTypeCol = "cellType",
                  imageIDCol = "imageID",
                  Rs = 1:10*10,
                  sigma = 50,
                  BPPARAM = BPPARAM)

topPairs(spicyTest, n = 10)
```

We can visualise these tests using `signifPlot` where we observe that cell type pairs appear to become less attractive (or avoid more) in the progression sample.

```{r}
# Visualise which relationships are changing the most.
signifPlot(
  spicyTest,
  breaks = c(-1.5, 1.5, 0.5)
)
```

`spicyR` also has functionality for plotting out individual pairwise relationships. We can first try look into whether the `SC` tumour cell type localises with the `GC` granular cell type, and whether this localisation affects progression vs non-progression of the tumour. 

```{r}
spicyBoxPlot(spicyTest, 
             from = "SC", 
             to = "GC")
```

Alternatively, we can look at the most differentially localised relationship between progressors and non-progressors by specifying `rank = 1`.

```{r}
spicyBoxPlot(spicyTest, 
             rank = 1)
```

## lisaClust: Find cellular neighbourhoods

[lisaClust package](https://www.bioconductor.org/packages/release/bioc/html/lisaClust.html)

`lisaClust` provides a series of functions to identify and visualise regions of tissue where spatial associations between cell-types is similar. This package can be used to provide a high-level summary of cell-type co-localisation in multiplexed imaging data that has been segmented at a single-cell resolution. Here we use the `lisaClust` function to clusters cells into 5 regions with distinct spatial ordering. 

```{r}
set.seed(51773)

cells <- lisaClust(
  cells,
  k = 4,
  sigma = 50,
  cellType = "cellType",
  BPPARAM = BPPARAM
)
```

### Region - cell type enrichment heatmap

We can interpret which spatial orderings the regions are quantifying using the `regionMap` function. This plots the frequency of each cell type in a region relative to what you would expect by chance. We can see here that our regions have neatly separated according to biological milieu, with region 1 and 2 representing our immune cell regions, region 3 representing our tumour cells, and region 4 representing our healthy epithelial and endothelial cells.

```{r}
# Visualise the enrichment of each cell type in each region
regionMap(cells, cellType = "cellType", limit = c(0.2, 2))
```

### Visualise regions

By default, these identified regions are stored in the `regions` column in the `colData` of our `SpatialExperiment` object. We can quickly examine the spatial arrangement of these regions using `ggplot` on image F3, where we can see the same division of immune, healthy, and tumour tissue that we identified in our `regionMap`.

```{r, message=FALSE, warning=FALSE}
cells |> 
  filter(imageID == "F3") |> 
  plotReducedDim("spatialCoords", colour_by = "region")
```

While much slower, we have also implemented a function for overlaying the region information as a hatching pattern so that the information can be viewed simultaneously with the cell type calls.

```{r}
# Use hatching to visualise regions and cell types.
hatchingPlot(
  cells,
  useImages = "F3",
  cellType = "cellType",
  nbp = 300
)
```

### Test for association with progression

Similar to cell type proportions, we can quickly use the `colTest` function to test for associations between the proportions of cells in each region and progression status by specifying `feature = "region"`.

```{r}
# Test if the proportion of each region is associated
# with progression status.
testRegion <- colTest(
  cells,
  feature = "region",
  condition = "group",
  type = "ttest")

testRegion
```

## Statial

[Statial
Package](https://www.bioconductor.org/packages/release/bioc/html/Statial.html)

### SpatioMark: Continuous changes in marker expression associated with varying levels of localisation.
The first step in analysing these changes is to calculate the spatial proximity (`getDistances`) of each cell to every cell type. These values will then be stored in the `reducedDims` slot of the `SingleCellExperiment` object under the names distances. SpatioMark also provides functionality to look into proximal cell abundance using the `getAbundance()` function, which is further explored within the `Statial` package vignette


```{r getDistances}
spe <- getDistances(spe,
  maxDist = 200,
  nCores = nCores,
  cellType = "cellType",
  spatialCoords = c("x", "y")
)
```

We can then visualise an example image, specified with `image = "F3"` and a particular marker interaction with cell type localisation. To visualise these changes, we specify two cell types with the `from` and `to` parameters, and a marker with the `marker` parameter (cell-cell-marker interactions). Here, we specify the changes in the marker podoplanin in `SC` as its localisation to `EP` increases or decreases, where we can observe that podoplanin decreases as its distance to the dense clump of tumour cells increases.


```{r plotStateChange}
p <- plotStateChanges(
  cells = spe,
  cellType = "cellType",
  spatialCoords = c("x", "y"),
  type = "distances",
  image = "F3",
  from = "SC",
  to = "EP",
  marker = "podoplanin",
  size = 1,
  shape = 19,
  interactive = FALSE,
  plotModelFit = FALSE,
  method = "lm"
)

p
```

SpatioMark aims to holistically uncover all such significant relationships by looking at all interactions across all images. The `calcStateChanges()` function provided by Statial can be expanded for this exact purpose - by not specifying cell types, a marker, or an image, `calcStateChanges()` will examine the most significant correlations between distance and marker expression across the entire dataset.

```{r stateChanges}
state_dist <- calcStateChanges(
  cells = spe,
  cellType = "cellType",
  type = "distances",
  assay = 2,
  nCores = nCores,
  minCells = 100
)

head(state_dist[state_dist$imageID == "F3",], n = 10)
```

The results from our SpatioMark outputs can be converted from a `data.frame` to a `matrix`, using the `prepMatrix()` function. Note, the choice of extracting either the t-statistic or the coefficient of the linear regression can be specified using the `column = "tval"` parameter, with the coefficient being the default extracted parameter. We can see that with SpatioMark, we get some features which are significant after adjusting for FDR.

```{r spatioMarkOutcome}
# Preparing outcome vector
outcome <- spe$group[!duplicated(spe$imageID)]
names(outcome) <- spe$imageID[!duplicated(spe$imageID)]

# Preparing features for Statial
distMat <- prepMatrix(state_dist)

distMat <- distMat[names(outcome), ]

# Remove some very small values
distMat <- distMat[, colMeans(abs(distMat) > 0.0001) > .8]

survivalResults <- colTest(distMat, outcome, type = "ttest")

head(survivalResults)
```

### Kontextual

[Pre-print](https://www.biorxiv.org/content/10.1101/2024.09.03.611109v1)

```{r hierarchy}
# Cluster cell type hierarchy
hierarchy <- treekoR::getClusterTree(t(assay(spe, "norm")),
                                     clusters = spe$cellType,
                                     hierarchy_method = "hopach")


parentList = Statial::getParentPhylo(hierarchy)

# Visualise tree
hierarchy$clust_tree |> 
  plot()

# Add parent 4 and 5
parentList$parent_4 = c(parentList$parent_2, parentList$parent_3)
parentList$parent_5 = c(parentList$parent_1, "EC")

all = unique(spe$cellType)

# Create data frame with all pairwise combinations of cells
treeDf = Statial::parentCombinations(all, parentList = parentList)

```

```{r kontextual, eval = FALSE}
kontext = Kontextual(
  cells = spe,
  cellType = "cellType",
  r = 50,
  parentDf = treeDf,
  cores = nCores
)
```

To save time we have saved the output of the results above.

```{r loadKontextual}
load(system.file("extdata", "kontextual.rda", package = "ScdneySpatial"))
```

```{r}
imageChoose = "H3"



kontext

spe |>
  colData() |>
  as.data.frame() |>
  filter(imageID == imageChoose) |>
  filter(cellType %in% c("SC", "TC_CD8", parentList$parent_4)) |>
  mutate(cellType = case_when(
    cellType %in% parentList$parent_4[parentList$parent_4 != "TC_CD8"] ~ "Immune",
    TRUE ~ cellType
  )) |> 
  mutate(cellType = factor(cellType, levels = c("TC_CD8",  "SC", "Immune"))) |> 
  arrange(desc(cellType)) |> 
  ggplot(aes(x = x, y = y, color = cellType)) +
  geom_point(size = 1) +
  scale_colour_manual(values = c("SC" = "#404040", "TC_CD8" = "#E25758", "Immune" = "#D7D8D8")) +
  guides(colour = guide_legend(title = "Cell types", override.aes = list(size = 3)))

```

We can use the `spicyR::spicy` to test for associations between the `Kontextual` values and progression status. To do so we use the `alternateResult` argument.

```{r kontextOutcome, fig.height = 5, fig.width = 10}
# Converting Kontextual result into data matrix
kontextMat <- prepMatrix(kontext)

# Replace NAs with 0
kontextMat[is.na(kontextMat)] <- 0

# Test association with outcome
kontextOutcome = spicyR::spicy(
  cells = spe,
  alternateResult = kontextMat,
  condition = "group",
  BPPARAM = BPPARAM
)

# Plot results
signifPlot(kontextOutcome)
```

# ClassifyR: Patient classification

[ClassifyR
Package](https://www.bioconductor.org/packages/release/bioc/html/ClassifyR.html)

[Journal
article](https://academic.oup.com/bioinformatics/article/31/11/1851/2365648?login=false)


Our ClassifyR package formalises a convenient framework for evaluating classification in R. We provide functionality to easily include four key modelling stages; Data transformation, feature selection, classifier training and prediction; into a cross-validation loop. Here we use the `crossValidate` function to perform 50 repeats of 5-fold cross-validation to evaluate the performance of a random forest applied to five quantifications of our IMC data; 1) Cell type proportions 2) Cell type localisation from `spicyR` 3) Region proportions from `lisaClust` 4) Cell type localisation in reference to a parent cell type from `Kontextual` 5) Cell changes in response to proximal changes from `SpatioMark`

```{r features}
# Create list to store data.frames
data <- list()

# Add proportions of each cell type in each image
data[["Proportions"]] <- getProp(spe, "cellType")

# Add pair-wise associations
# spicyMat <- bind(spicyTest)
# spicyMat[is.na(spicyMat)] <- 0
# spicyMat <- spicyMat |>
#   select(!condition) |>
#   tibble::column_to_rownames("imageID")
# 
# data[["SpicyR"]] <- spicyMat

# Add proportions of each region in each image to the list of dataframes.
#data[["LisaClust"]] <- getProp(cells, "region")


# Add SpatioMark features
#data[["SpatioMark"]] <- distMat

# Add Kontextual features
data[["Kontextual"]] <- kontextMat
```


```{r classification}
# Set seed
set.seed(51773)

# Perform 5-fold cross-validation with 50 repeats.
cv <- crossValidate(
  measurements = data,
  outcome = outcome,
  selectionMethod = "t-test",
  classifier = "LASSOGLM",
  nFolds = 5,
  nFeatures = 50,
  nRepeats = 50,
  nCores = nCores
)
```

## Visualise cross-validated prediction performance

Here we use the `performancePlot` function to assess the AUC from each repeat of the 5-fold cross-validation.

```{r performancePlot}
performancePlot(
  cv,
  metric = "AUC",
  characteristicsList = list(x = "Assay Name"),
  orderingList = list("Assay Name" = c("Proportions", "SpicyR", "LisaClust", "Kontextual", "SpatioMark"))
)
```


::: question
**Questions**

1. Run the following chunk of code, what is this plot telling us?
2. Do different methods perform differently on different samples?
3. Knowing this, what might we do to improve our model?
:::



```{r sampleMetrics}
samplesMetricMap(cv) |> 
  plot()

```


```{r multiViewClassification}
cv <- crossValidate(
  measurements = data,
  outcome = outcome,
  classifier = "randomForest",
  nFolds = 5,
  nRepeats = 50,
  multiViewMethod = "merge",
  nCores = nCores
)

performancePlot(
  cv,
  metric = "AUC",
  characteristicsList = list(x = "Assay Name"),
)
```